---
title: 'Solar panel analysis pt 3: Scanning for objects'
author: Jeroen Boeye
date: '2017-11-01'
slug: solar-panel-analysis-pt-3-Scanning-for-objects
comments: true
categories: [R]
tags: [Solar, Photovoltaic, R]
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, 
                      echo = TRUE, dpi = 180)
options(width=80, dplyr.width = 150)

```

### Intro

In the first posts on my solar panel data-set we did some [exploration](https://www.jeroenboeye.com/blog/solar-panel-analysis-pt-1-exploration/) and found a number of interesting patterns. One of these was the effect of temperature on the efficiency of the panels which we investigated in the [second post](https://www.jeroenboeye.com/blog/solar-panel-analysis-pt-2-temperature-and-efficiency/). Another interesting pattern was that for a number of months the maximal power throughout the day did not follow the expected bell shape curve. Let's load the [data](https://github.com/jeroenboeye/jbo_blog/blob/master/static/files/solar_power.csv) and visualize this for June and December so you get an idea of what I'm talking about.

```{r, fig.width=6, fig.height=4}
library(rprojroot)
library(tidyverse)
library(lubridate)
rt <- rprojroot::is_rstudio_project

solarplant_df <- read_csv2(rt$find_file("static/files/solar_power.csv"))%>%
  # the times in the raw dataset have CET/CEST timezones
  # by default they will be read as UTC
  # we undo this by forcing the correct timezone 
  # (without changing the times)
  mutate(timestamp = force_tz(timestamp, tzone = 'CET'))%>%
  # Since we don't want summer vs winter hour differences
  # we now transform to UTC and add one hour. 
  # This puts all hours in CET winter time
  mutate(timestamp = with_tz(timestamp, tzone = 'UTC') + 3600)%>%
  # temperature not selected for this post
  select(timestamp, power)%>%
  # round the timestamps to 10 minutes.
  # +95% of the data is allready in this format
  mutate(timestamp = round_date(timestamp, unit = "10 minutes"))

solarplant_df%>%
  mutate(hour = hour(timestamp) + minute(timestamp) / 60,
         month = month(timestamp, label = T, abbr = F))%>%
  filter(month %in% c('December', 'June'))%>%
  select(month, hour, power)%>%
  group_by(month, hour)%>%
  mutate(power = max(power))%>%
  ggplot(aes(x = hour, y = power)) +
  geom_point(alpha = 0.8, size = rel(0.8), shape = 21) +
  facet_wrap(~month) +
  labs(x = 'Hour of the day',
       y = 'Power (W)') +
  theme_minimal() +
  theme(text = element_text(colour = "grey40"),
        strip.text.x = element_text(colour = "grey40"),
        panel.border = element_rect(colour = "grey40", 
                                    size = rel(0.3),
                                    fill = NA),
        legend.position = 'top',
        legend.justification = 'left')
```

Looking at the highest values for time of the day we get an asymmetrical shape with lower maximum values in the morning in June while we get less than expected maximum values around noon in December. This pattern is caused by two trees to the east and south of my home which cast a shade on the roof during different times of the day. The tree to the south is not too close to the house and only when the Sun's elevation is very low in winter does it's shade reach the roof.


### Visualising the obscuring objects

While I know that the trees are behind this pattern in the data I was wondering if we could derive their location and shape from the data. If we could somehow get a value for the power loss for each timestamp in the data set we should be able to couple this data to a position of the Sun at that time using the  **`solarpos`** function from the **`maptools`** package. Then we could create a panoramic picture with each position of the sun throughout the year as seen by the panels and highlight Solar positions that consistently yield less power and are thus obscured by an object.
This sounds like a cool idea but to get there we first need to get an idea of the power loss per timestamp. We'll have to create a model that gives us the theoretical maximum output for each timestamp and then calculate the difference with the observed values.

### Visualising a full year of solar positions

Before we dive into modelling the maximal power outputs let's look at the solar positions we will bind the power loss data to. We get all solar positions for a full year in 10 minute intervals and filter out the values where the Sun is below the horizon (negative elevation).
```{r, fig.width=7, fig.height=5}
library(sp)
library(maptools)
library(viridis)

# My home coordinates, needed to calculate the solar elevation
home_coords <- sp::SpatialPoints(matrix(c(4.401168, 51.220305), nrow=1), 
                                 proj4string=sp::CRS("+proj=longlat +datum=WGS84"))


# create a dataframe with 10 minute time intervals for a whole year
solar_position_df <- tibble(timestamp = seq(from = ymd_hms("1970-01-01 00:00:00", tz = "UTC"),
                                            to = ymd_hms("1970-12-31 23:50:00", tz = "UTC"),
                                            by = '10 min'))%>%
  # calculate the solar position for each timestamp at my home coordinates
  # the solarpos function from the maptools package returns a list with 2 elements
  # 1 = the solar azimuth angle (degrees from North), 2 = the solar altitude
  mutate(solar_azimuth_angle = maptools::solarpos(home_coords, timestamp)[,1],
         solar_elevation = maptools::solarpos(home_coords, timestamp)[,2])%>%
  # we only care about solar elevations > 0
  filter(solar_elevation > 0)%>%
  # The solar panel dataframe uses UTC + 1h 
  # so lets set this data up the same way
  mutate(timestamp = timestamp + 3600)

# visualise the position of the sun per hour and give each month a colour
solar_position_df%>%
  # only visualise values on round hours
  filter(minute(timestamp) == 0)%>%
  ggplot(aes(x = solar_azimuth_angle, 
             y = solar_elevation, 
             colour = month(timestamp, label = T, abbr = F)))+
  geom_point(alpha = 0.5)+
  scale_x_continuous(breaks = seq(45, 315, 45), 
                     labels = c("Northwest", "West", 
                                "Southwest", "South", 
                                "Southeast", "East", 
                                "Northeast"))+
  viridis::scale_colour_viridis(discrete = T)+
  scale_y_continuous(expand = c(0, 0))+
  labs(y = "Solar elevation (°)")+
  theme_minimal()+
  theme(strip.text.y = element_text(angle = 0, hjust = 0),
        axis.title.x = element_blank(),
        legend.title = element_blank(),
        panel.grid = element_blank(),
        panel.background = element_rect(fill = NA, colour = "grey40"))
```
These are the Sun's positions throughout the year on each round hour. The resulting shapes are called [analemmas](https://en.wikipedia.org/wiki/Analemma) and they are really interesting. Back in ancient times they used to give astronomers headaches since they cause small deviations in the time read from sundials throughout the year. The issue was eventually resolved when they came up with the [equation of time](https://en.wikipedia.org/wiki/Equation_of_time). I recommend a dive down the good old Wikipedia rabbit hole on this topic but for now, let's get back to our solar panel data!

### Modelling maximal power output

We want a model that produces nice, symmetrical, bell shaped curves that are bigger in summer and smaller in winter. To do this we have to think about what is actually causing this variation in the power yield. In winter the Sun transfers less energy to the panels because of it's lower elevation but what exactly causes this lower elevation to be less efficient? 

#### Energy loss due to spreading

Image you carry a flashlight in a dark room. When you point it directly at the wall you get a small circle that is intensely lit. However, if you stand close to the wall and point the flashlight somewhat sideways you light a bigger surface. The energy emitted by the flashlight is spread out and each square centimeter in the lit area now receives less energy. We can calculate this effect with cos(pi / 2 - angle). When we transform a 90° angle to radials we get  pi / 2 if we enter this in the function we get the cosine of zero which is 1 or 100% spreading efficiency. In Belgium the Sun has a maximal elevation of 62° which has a spreading efficiency of 88%. During the winter solstice the Sun only reaches 15.3° at noon which has a spreading efficiency of only 26% so that's almost 3.5 fold reduction in solar power on a flat surface due to the changing vertical position of the sun. Fortunately, the panels themselves are somewhat inclined (20%) towards the South so that even in winter the angle of arrival is closer to 35° (57% efficiency). However, this inclination causes a more complex spreading effect because the panels now point in a particular direction. I have estimated both effects in the code below.

```{r}


# angles in degrees to angles in radials
# the '_deg' part of the variable name shows it's in degrees
deg_to_rad <- function(angle_deg){pi * angle_deg / 180}

# vertical spreading effect
get_vertical_spreading <- function(angle_deg){
  cos(pi / 2 - deg_to_rad(angle_deg))
  }

# horizontal spreading effect 
# because the panels point towards the South
get_horizontal_spreading <- function(angle_deg){
  ((-cos(deg_to_rad(angle_deg)) + 1) / 2) * 0.3 + 0.7
  }
```


#### Energy loss due to traveling through the atmosphere

A second, less straightforward, effect that decreases the energy transfer from the Sun to the panels at lower Solar elevations is the fact that the distance traveled through the atmosphere becomes larger. As the light rays make their way through this mass of air they often collide with the molecules and become scattered. It is this [effect](https://en.wikipedia.org/wiki/Rayleigh_scattering) that makes sunsets so beautiful and makes the sky above above us appear blue instead of black. 
Some clever people have studied this effect in great detail and we can take their [equations](https://en.wikipedia.org/wiki/Air_mass_(solar_energy)#math_I.1) from Wikipedia to learn that at a 90° angle the energy per square meter is about 1051 Watts. At 62° (Belgium summer solstice) we get 1019 Watts/m² which is still quite good. However at 15.3° in winter we fall back to 628 Watts/m² which is a 38% reduction compared to the energy per surface in summer.


```{r}

# calulate Air Mass (AM) according to Karsten & Young 
# https://en.wikipedia.org/wiki/Air_mass_(solar_energy)#math_I.1
# https://en.wikipedia.org/wiki/Solar_irradiance
get_air_mass <- function(angle_deg){
  1 / (cos(deg_to_rad(angle_deg)) + 0.50572 * (96.07995 - angle_deg) ^ -1.6364)
}
 
# this function calculates how much power (W/m²?) 
# is left in sun rays reaching the surface
# the only effect weakening solar power is air mass at this point
# incoming solar power on top of atmosphere = 1366 W/m?
get_solar_intensity <- function(angle_deg){
  # extract from 90 to get correct z angle (see wikipedia)
  1.1 * 1366 * 0.7 ^ get_air_mass(90 - angle_deg) ^ 0.678
  }
```

#### Putting it all together

Now that we have created functions for the most important effects at play it's time to put them together and create a new function that will put them all together and estimate the maximal power yield of the solar installation when we feed it the position of the sun.
```{r}
solar_position_to_max_power <- function(angle_deg, 
                                        azimuth_angle_deg){
  
  # how much of the solar intensity is left after travelling through the air mass.
  solar_intensity <- get_solar_intensity(angle_deg)
  
  # light spreading out because of sideways spreading 
  horizontal_spreading <- get_horizontal_spreading(azimuth_angle_deg)
  
  # panels are inclined at 20°
  vertical_spreading <- get_vertical_spreading(angle_deg + 20)
  
  surface_power <- solar_intensity * vertical_spreading * horizontal_spreading
  
  # 7 panels with 1.6 m²
  panel_surface <- 1.6 * 7 
  
  # 16% efficiency of panels
  efficieny <- 0.16
  
  # eventual power is the panel surface times 
  # the power left in the Sun's rays on the panel times
  # the efficiency of the panels
  max_plant_power <- panel_surface * surface_power * efficieny 
  
  return(max_plant_power)
}
```

We can now call this function on the data set with solar positions for a full year which we created earlier on.

```{r}
solar_position_max_power_df <- solar_position_df%>%
  # calculate the maximal power of the solar panels per timestamp
  # we need this to check if the maximal observed power is much lower 
  # than the maximal theoretical power
  mutate(max_power = solar_position_to_max_power(solar_elevation, 
                                                 solar_azimuth_angle))


```

Now let's bring both datasets together and calculate the ratio between the power observed and the estimated maximal power.

```{r}
combined_df <- solarplant_df%>%
  # add the solar plant data to the solar position data
  # standardise the time to 1970 to allow joing the solar position dataset
  mutate(timestamp_basic = lubridate::origin + (yday(timestamp) - 1) * 3600 * 24 + 
                           hour(timestamp) * 3600 + minute(timestamp) * 60)%>%
  right_join(solar_position_max_power_df, by = c("timestamp_basic" = "timestamp"))%>%
  replace_na(list(power = 0))%>%
  # efficiency is power / max power but can't be higher than 1 (100%)
  mutate(efficiency = ifelse(power / max_power > 1, 1, power / max_power),
         # some time variables we need later on
         hour = hour(timestamp) + minute(timestamp) / 60,
         week_date = round_date(timestamp_basic, unit = "week"))
```

To check if our theoretical model gives acceptable results let's visualise both observed and estimated maximal power for a some sampled days.
```{r, fig.width=6, fig.height=4}
combined_df%>%
  mutate(date = as_date(timestamp))%>%
  filter(date %in% as_date(c("2017-06-01", "2016-12-04")))%>%
  # filter(date %in% seq(from = as_date("2016-12-01"),
  #                      to = as_date("2016-12-20"),
  #                      by = "day"))%>%
  ggplot(aes(x = hour(timestamp) + minute(timestamp) / 60, y = max_power))+
  geom_line()+
  geom_point(aes(y = power))+
  facet_wrap(~date)


```
As you can see our theoretical estimation fits the observed values quite well both in winter and summer. This should now allow us to detect solar positions consistently yield lower than maximal values.

#### Visualizing maximal efficieny per solar position

Most of the time when the panels yield less than what is theoretically possible, clouds are spoiling the party. To filter out this noise and let the effect of stationary objects emerge we will look at the maximal efficiency for each time of the day, per week of the year. Since we have almost 2 years in the dataset we now have registered 14 days for most weeks of the year. For clouds to appear as a stationary object in our analysis it would have to be cloudy on a particular time of the day for all 14 days measured. Even for Belgium this is rarely the case.

```{r, fig.width=10, fig.height=6}
max_efficiency_df <- combined_df%>%
  # for each week of the year we hope the sun shined at least once per 10 minute interval
  # this way we can get an accurate maximal efficiency estimate by grouping 
  # on the week and hour variables and calculating the maximal values for all relevant variables
  group_by(week_date, hour)%>%
  summarise(solar_azimuth_angle = max(solar_azimuth_angle),
            solar_elevation = max(solar_elevation),
            efficiency = max(efficiency))%>%
  ungroup()

max_efficiency_df%>%
  # only show points where the maximal efficiency is below 50%
  #filter(efficiency < 0.5)%>%
  ggplot(aes(x = solar_azimuth_angle, 
             y = solar_elevation, 
             colour = efficiency, 
             alpha = -efficiency, 
             size = -efficiency))+
  #annotation_custom(g, xmin = -50, xmax = 320, ymin = -15, ymax = 70)+
  geom_point()+
  scale_color_gradient2(low = "darkgreen", mid = "lightblue", high = "lightblue", midpoint = 0.7)+
  scale_alpha_continuous(range = c(0.2, 1))+
  scale_x_continuous(breaks = seq(45, 315, 45), 
                     labels = c("Northwest", "West", "Southwest", "South", "Southeast", "East", "Northeast"))+
  scale_y_continuous(expand = c(0, 0), limits = c(0, 65))+
  labs(y = "Solar elevation (°)")+
  theme(panel.background = element_blank(),#element_rect(fill = "lightblue"),
        panel.grid = element_blank(),
        axis.title.x = element_blank())
```
As you can see I've taken some artistic liberty in creating this plot. Low maximal efficiencies are coloured in green (as I know they are mostly caused by trees) and high values maximal values get a blue colour + higher transparency and smaller circles.

To verify this data I got up on my roof and created an actual panoramic picture. Let's use it as the background for our plot.

```{r}
library(jpeg)
library(grid)
img <- jpeg::readJPEG(rt$find_file("static/figs/roof_panorama.jpg"))
#img <- readPNG("data/panorama.png") 
g <- rasterGrob(img, interpolate = TRUE) 

max_efficiency_df%>%
  # only show points where the maximal efficiency is below 50%
  #filter(efficiency < 0.5)%>%
  ggplot(aes(x = solar_azimuth_angle, 
             y = solar_elevation, 
             colour = efficiency, 
             alpha = -efficiency, 
             size = -efficiency))+
  annotation_custom(g, xmin = -50, xmax = 320, ymin = -15, ymax = 70)+
  geom_point()+
  scale_color_gradient2(low = "darkgreen", mid = "lightblue", high = "lightblue", midpoint = 0.7)+
  scale_alpha_continuous(range = c(0.2, 1))+
  scale_x_continuous(breaks = seq(45, 315, 45), 
                     labels = c("Northwest", "West", "Southwest", "South", "Southeast", "East", "Northeast"))+
  scale_y_continuous(expand = c(0, 0), limits = c(-10, 70))+
  theme(panel.background = element_blank(),
        panel.grid = element_blank())
```

